---
title: "Introduction to pls4all"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to pls4all}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`pls4all` is a portable Partial Least Squares (PLS) and Near-Infrared
Spectroscopy (NIRS) engine. The C++17 implementation is vendored and
compiled from source at install time; no external system library is
required.

## Quick start

```{r quickstart}
library(pls4all)

set.seed(42)
n <- 50
p <- 8
X <- matrix(rnorm(n * p), nrow = n, ncol = p)
beta <- c(2, -1, 0.5, rep(0, p - 3))
y <- as.numeric(X %*% beta + rnorm(n, sd = 0.1))

fit <- pls(X, y, n_components = 3)
preds <- predict(fit, newdata = X)

cat("RMSE:", sqrt(mean((preds - y)^2)), "\n")
```

## Version + ABI

The runtime version string embeds both the project semver and the
stable C ABI version. This is useful for downstream consumers that
need to assert ABI compatibility before invoking lower-level
entry points.

```{r version}
pls4all_version()
pls4all_abi_version()
```

## Supported algorithms

`pls()` selects the solver via the `algo` argument. Recognised values
include `"pls_nipals"` (default), `"pls_simpls"`, `"pls_svd"`,
`"pls_kernel_algorithm"`, `"pls_wide_kernel"`,
`"pls_orthogonal_scores"`, `"pls_power"`, `"pls_randomized_svd"`, and
`"pcr_svd"`. The full algorithm taxonomy is documented in the package
website.

```{r algos}
fit_simpls <- pls(X, y, n_components = 3, algo = "pls_simpls")
fit_svd    <- pls(X, y, n_components = 3, algo = "pls_svd")

# Same numerical answers up to algorithmic tolerance.
all.equal(predict(fit_simpls, newdata = X[1:5, ]),
          predict(fit_svd,    newdata = X[1:5, ]),
          tolerance = 1e-10)
```

## Further reading

* The `dispatch_fit()` entry point exposes the full method catalogue
  (sparse SIMPLS, CPPLS, weighted, robust, ridge, continuum,
  multi-block, GLM, MIR, PDS, DS, and others) with a unified
  parameter-list interface.
* Variable-selection wrappers (`spa_select`, `cars_select`,
  `variable_select_rank`) return a ranked feature index suitable for
  downstream model retraining.
* Diagnostics (`pls_diagnostics_compute`, `approximate_press_compute`,
  `pls_monitoring_run`) implement Hotelling T², Q residuals, DModX,
  and process-monitoring alarms.

See the project [website](https://gbeurier.github.io/pls4all/) and the
cross-binding parity reports at
<https://github.com/GBeurier/pls4all/tree/main/docs/parity> for the
full feature matrix.
