---
title: "Introduction to n4m"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to n4m}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`n4m` is a portable Partial Least Squares (PLS) and Near-Infrared
Spectroscopy (NIRS) engine. The C++17 implementation is vendored and
compiled from source at install time; no external system library is
required.

## Quick start (matrix API)

```{r quickstart}
library(n4m)

set.seed(42)
n <- 50
p <- 8
X <- matrix(rnorm(n * p), nrow = n, ncol = p)
beta <- c(2, -1, 0.5, rep(0, p - 3))
y <- matrix(as.numeric(X %*% beta + rnorm(n, sd = 0.1)), ncol = 1)

fit <- n4m_fit(X, y, algo = "pls_nipals", n_components = 3L)
preds <- n4m_predict(fit, X)

cat("RMSE:", sqrt(mean((preds - y)^2)), "\n")
```

## Formula interface

The base-R formula API mirrors `lm()` and other standard regression
modelling conventions.

```{r formula}
df <- as.data.frame(X)
df$y <- as.numeric(y)
fit_form <- pls(y ~ ., data = df, ncomp = 3L)
class(fit_form)
```

## Version and ABI

The runtime version string embeds both the project semver and the
stable C ABI version. Downstream consumers can assert ABI
compatibility before calling lower-level entry points.

```{r version}
n4m_version()
n4m_abi_version()
```

## Supported algorithms

`n4m_fit()` selects the solver via the `algo` argument. Recognised
values include `"pls_nipals"` (default), `"pls_simpls"`, `"pls_svd"`,
`"pls_kernel_algorithm"`, `"pls_wide_kernel"`,
`"pls_orthogonal_scores"`, `"pls_power"`, `"pls_randomized_svd"`, and
`"pcr_svd"`. The full algorithm taxonomy is documented in the package
website.

```{r algos}
fit_simpls <- n4m_fit(X, y, algo = "pls_simpls", n_components = 3L)
fit_svd    <- n4m_fit(X, y, algo = "pls_svd",    n_components = 3L)

# Same numerical answers up to algorithmic tolerance.
all.equal(n4m_predict(fit_simpls, X[1:5, ]),
          n4m_predict(fit_svd,    X[1:5, ]),
          tolerance = 1e-10)
```

## Further reading

* The `dispatch_fit()` entry point exposes the full method catalogue
  (sparse SIMPLS, CPPLS, weighted, robust, ridge, continuum,
  multi-block, GLM, MIR, PDS, DS, and others) with a unified
  parameter-list interface.
* Variable-selection wrappers (`spa_select`, `cars_select`,
  `variable_select_rank`) return a ranked feature index suitable for
  downstream model retraining.
* Diagnostics (`pls_diagnostics_compute`, `approximate_press_compute`,
  `pls_monitoring_run`) implement Hotelling T-squared, Q residuals,
  DModX, and process-monitoring alarms.

See the project [repository](https://github.com/GBeurier/nirs4all-methods) and the
cross-binding parity reports at
<https://github.com/GBeurier/nirs4all-methods/tree/main/docs/parity> for the
full feature matrix.
